Convergence rate of MAP estimates \\ for the exponential family

### 2021-10-22 aistats.tex
Be more gentle on the covariance
Quick typo pass on Appendix + clarify erratum. RESUBMITTED

\addcontentsline{toc}{section}{Appendix} % Add the appendix text to the document TOC
\part{Appendix} % Start the appendix part
\begin{important}
	{\bf Erratum:}  contrary to what is written in the main text, the log-partition of a Gaussian \emph{is} self-concordant. See \cref{app:gaussian} for details. 
\end{important}
\parttoc % Insert the appendix TOC

@@

\begin{important}
	{\bf Erratum:}  
	In the main text, we claimed that the log-partition of a Gaussian was not self-concordant. 
	There was an error in our derivations and we will now prove the opposite.
	However this does not change our story, and in particular our concern about the lack of results in the few sample regime remains valid.
	%TODO ostroskii and bach are using a subgaussian assumption which does not hold for a Gaussian, so there might still be work to do. But also check anastasiou to be sure.
\end{important}

@@

The Gaussian log-partition function and entropy are, up to constants,
\alignn{
	\logpart(\nat) &= \frac{\nat_1^2}{-4\nat_2} - \half \log(-\nat_2) \\
	\conj(\m) &= - \half \log (\mu_2 - \mu_1^2) \; ,
}
where $\nat_1\in\real, \nat_2 <0$ and $\mu_1\in\real, \mu_2>\mu_1^2$.
% TODO These functions and their domains are represented in Figure \TODO{}.
We provide definitions of self-concordance in \cref{app:self-concordant}.

\paragraph{Entropy is Self-Concordant,}
\citet[Example 4.1.1.4,   p.177]{nesterov2003introductory} proves that logarithmic barriers for second order regions are self-concordant, 
that is functions of the form 
\alignn{
	f(\theta) = -\log\paren{\alpha + \lin{a, \theta} - \frac{1}{2}\lin{\mA \theta, \theta}}
	\text{ on }
	\left\{\theta\in\mathbb{R}^n \cond \alpha + \lin{a, \theta} - \frac{1}{2}\lin{\mA \theta, \theta} > 0\right\}.
}
The entropy fits into this definition with $\mA = \begin{pmatrix} 2 & 0 \\ 0 & 0 \end{pmatrix}$ and $a = ( 0 \ 1 )^\top$.

\paragraph{Log-partition is Self-Concordant}
As proved in \citet{nesterov1994interior} self-concordance is preserved by Fenchel conjugacy.
Since $\conj$ is self-concordant, $\logpart$ as well.
For a more accessible reference, see also \citet[Prop.~6]{sun2019generalized}.

@@

In this section, we define self-concordance, we prove proposition~\ref{prop:selfConcordant}, and we prove that a 1 dimensional function is self-concordant if and only if its convex conjugate is self-concordant as well.

@@

{\bf Correction:} In the main text, we claimed that the Fenchel conjugate of a 1 dimensional function is also self-conjugate. 
Actually, this is also true in higher dimension, as proved by \citet{nesterov1994interior}. See \citet[Prop.~6]{sun2019generalized} for a more accessible reference.

@@

We quickly reviews some important properties of self-concordant functions, introduced in \citep{nesterov2003introductory}. We start with a few notations. Let $A^*$ be a self-concordant function. Then, we write
\begin{itemize}
	\item the local norm $\|\cdot\|_x = \sqrt{ \langle \nabla^2A^*(x)\cdot,\, \cdot \rangle }$ 
	\item the distance function $\omega(t) = t-\ln (1+t),\;t\geq 0$, and its dual $\;\omega^*(t) = -t-\ln(1-t)$ defined for $t\in [0,1]$.
\end{itemize}
Note that $\omega^*(t)$ is positive, convex and monotonically increasing for $t\in[0,1]$.

@@

The next result show that, if $y$ is sufficiently close to $x$, then we can bound the Bregman divergence of $A^*$ using the distance function $\omega^*$ and local norms.
\begin{proposition}(Upper bound of self-concordant functions \citep[Theorem 4.1.8]{nesterov2003introductory}) \label{prop:upper_bound_self_concordance}
	For any $x,\,y\in\dom A^*$, if $\|y-x\|_x<1$, then
	\[
		\bregmanconj(y,x) \leq \omega^*(\|y-x\|_x).
	\]
\end{proposition}

@@

\subsection{Expectation of SMD's Variance Assumption}
The first step is to notice the symmetrized Bregman can be expressed as an inner product between primal and dual parameters
\alignn{
	\cS_\conj (\mu , \bar\mu)
	= \lin{ \nabla \conj(\mu) - \nabla \conj(\bar\mu) , \mu - \bar\mu}
	= \lin{ \nat -\bar\nat , \mu - \bar\mu} \; .
}
Now notice that~\eqref{eq:hanzely} features $\mu := \hat \mu_{n+1} = \hat\m_{n} - \lr g(\hat \nat_{n})$ stochastic and $\bar\mu := \bar\mu_{n+1} =  \hat\m_{n}   -\lr \nabla f(\hat \nat_{n})$ deterministic such that $\E_g[\mu] = \bar\mu$.
This means that
\alignn{
	\expect{\cS_\conj (\mu , \E[\mu])} 
	= \expect{\lin{\nat, \mu - \E[\mu]}}
	=\Cov(\nat, \mu)
	=\expect{\lin{\nat - \E[\nat],\mu}} 
	= \expect{\cS_\logpart(\nat, \E[\nat])} \; .
}
where the last equality holds by symmetry between the roles of $\logpart$ and $\conj$.
Note that the middle covariance formulation is actually the one  used by \citet{hanzely2018fastest}.
Now, \cref{eq:bias-variance} defines the variance as 
\alignn{
	\expect[1:n]{\bregmanconj(\tilde \m_n , \hat\m_n) }
	=\expect[1:n]{\bregman(\hat\nat_n , \E_{1:n}[\hat\nat_n]) }
	\leq \expect[1:n]{\cS_A(\hat\nat_n , \E_{1:n}[\hat\nat_n]) }
	=  \expect[1:n]{\cS_\conj(\hat\mu_n , \E_{1:n}[\hat\mu_n]) }\;,
	\label{eq:bregman-covariance}
}
where expectations $\E_{1:n}$ are on all samples $X_1, \dots, X_n$, whereas~\eqref{eq:hanzely} is written with 
\alignn{
	\expect[n]{\cS_\conj(\hat\mu_n , \E_{n}[\hat\mu_n]) } \leq \lr^2 C \; ,
}
where the expectation $\E_{n}$ is taken over only the last sample $X_n$, and the bound should hold uniformly over all $\hat\mu_{n-1}$.
Taking the expectation over $\hat\mu_{n-1}$ instead yields
\alignn{
	\expect[1:n]{\cS_\conj(\hat\mu_n , \E_{n}[\hat\mu_n]) }  \leq \lr^2 C \;.
}
The only difference with the RHS of \cref{eq:bregman-covariance} is in the inner expectation.
To overcome this difference, we need to plug in the form of $\hat \mu_n = \frac{n_0\mu_0 + \sum_i T_i}{n_0 +n}$. Notice that 
\alignn{
	&\MAPm - \E_n[\MAPm] = \frac{T_n - \m^*}{n_0+n} \\
	\implies & \expect[1:n]{\cS_\conj(\hat\mu_n , \E_{n}[\hat\mu_n]) }
	= \inv{n_0+n} \expect[1:n]{ \lin{\MAPt , T_1 - \m^*} }
}
while
\alignn{
	&\MAPm - \E_{1:n}[\MAPm] = \frac{\smallsum_i (T_i - \m^*)}{n_0+n} \\
	\implies & \expect[1:n]{\cS_\conj(\hat\mu_n , \E_{1:n}[\hat\mu_n]) }
	=\frac{n}{n_0+n} \expect[1:n]{ \lin{\MAPt , T_1 - \m^*} } \; .
}
In the end we get that the variance is dominated by $n$ times the expectation of \cref{eq:hanzely}
\alignn{
	\expect[1:n]{\bregmanconj(\tilde \m_n , \hat\m_n) }
	\leq n \expect[1:n]{\cS_\conj(\hat\mu_n , \E_{n}[\hat\mu_n]) }  \; .	
}
If assumption~\eqref{eq:hanzely} holds, then
\alignn{
\expect[1:n]{\bregmanconj(\tilde \m_n , \hat\m_n) }
\leq n \lr_n^2 C \in O\paren{\inv{n}} \; ,
}
where we assumed $\lr_n  \in O\paren{\inv{n}}$.
In conclusion, assuming~\eqref{eq:hanzely} holds uniformly or in expectation immediately implies a $O\paren{\inv{n}}$ convergence rate on the variance.

>>>>

\addcontentsline{toc}{section}{Appendix} % Add the appendix text to the document TOC
\part{Appendix} % Start the appendix part
\begin{important}
	{\bf Erratum:} contrary to what is written in the main text in the last paragraph of Section~\ref{ssec:gaussian} and~\ref{ssec:local-quadratic}, the log-partition of a full Gaussian \emph{is} self-concordant. See \cref{app:gaussian} for details. This does not change the narrative of the paper though.
\end{important}
\parttoc % Insert the appendix TOC

@@

\begin{important}
	{\bf Erratum:}  
	In the main text, we claimed by mistake in the last paragraph of Section~\ref{ssec:gaussian} and~\ref{ssec:local-quadratic} that the log-partition of a full Gaussian is not self-concordant. 
	There was an error in our derivations and we will now prove the opposite (using classical results).
	However, this does not change our main story, and in particular, the point that we made in the last paragraph of Section~\ref{ssec:local-quadratic} that there is a lack of results in the few sample regime remains valid.
	%TODO ostroskii and bach are using a subgaussian assumption which does not hold for a Gaussian, so there might still be work to do. But also check anastasiou to be sure.
\end{important}

@@

The Gaussian log-partition function and entropy are, up to constants,
\alignn{
	\logpart(\nat) &= \frac{\nat_1^2}{-4\nat_2} - \half \log(-\nat_2) \\
	\conj(\m) &= - \half \log (\mu_2 - \mu_1^2) \; , \label{eq:app-entropy}
}
where $\nat_1\in\real, \nat_2 <0$ and $\mu_1\in\real, \mu_2>\mu_1^2$.
% TODO These functions and their domains are represented in Figure \TODO{}.
We provide definitions of self-concordance in \cref{app:self-concordant}.

\paragraph{Entropy is Self-Concordant.}
\citet[Example 4.1.1.4,   p.177]{nesterov2003introductory} proves that logarithmic barriers for second order regions are self-concordant, 
that is functions of the form 
\alignn{
	f(\theta) = -\log\paren{\alpha + \lin{a, \theta} - \frac{1}{2}\lin{\mA \theta, \theta}}
	\text{ on }
	\left\{\theta\in\mathbb{R}^n \cond \alpha + \lin{a, \theta} - \frac{1}{2}\lin{\mA \theta, \theta} > 0\right\}.
}
The entropy~\eqref{eq:app-entropy} fits into this definition with $\mA = \begin{pmatrix} 2 & 0 \\ 0 & 0 \end{pmatrix}$ and $a = ( 0 \ 1 )^\top$.

\paragraph{Log-partition is Self-Concordant.}
As proved in \citet{nesterov1994interior}, self-concordance is preserved by Fenchel conjugacy.
Since $\conj$ is self-concordant, $\logpart$ is as well.
For a more accessible reference, see also \citet[Prop.~6]{sun2019generalized}.

@@

In this section, we define self-concordance and we prove Proposition~\ref{prop:selfConcordant}.

@@

{\bf Clarification:} In the main text in Section~\ref{ssec:local-quadratic}, we claimed that the Fenchel conjugate of a 1-dimensional function is also self-concordant. 
Actually, this is also true in higher dimensions, as proved by \citet{nesterov1994interior}. See \citet[Prop.~6]{sun2019generalized} for a more accessible reference.

@@

We quickly review some important properties of self-concordant functions, introduced in \citep{nesterov2003introductory}. We start with a some notation. Let $A^*$ be a self-concordant function. Then, we write
\begin{itemize}
	\item the local norm $\|\cdot\|_x = \sqrt{ \langle \nabla^2A^*(x)\cdot,\, \cdot \rangle }$ 
	\item the distance function $\omega(t) = t-\ln (1+t),\;t\geq 0$, and its dual $\;\omega^*(t) = -t-\ln(1-t)$ defined for $t\in [0,1]$.
\end{itemize}
Note that $\omega^*(t)$ is positive, convex and monotonically increasing for $t\in[0,1]$.

@@

The next result shows that, if $y$ is sufficiently close to $x$, then we can bound the Bregman divergence of $A^*$ using the distance function $\omega^*$ and local norms.
\begin{proposition}(Upper bound of self-concordant functions \citep[Theorem 4.1.8]{nesterov2003introductory}) \label{prop:upper_bound_self_concordance}
	For any $x,\,y\in\dom A^*$, if $\|y-x\|_x<1$, then
	\[
		\bregmanconj(y,x) \leq \omega^*(\|y-x\|_x).
	\]
\end{proposition}

@@

\subsection{Expectation of SMD's Variance Assumption}
The first step is to notice the symmetrized Bregman can be expressed as an inner product between primal and dual parameters
\alignn{
	\cS_\conj (\mu , \bar\mu)
	= \lin{ \nabla \conj(\mu) - \nabla \conj(\bar\mu) , \mu - \bar\mu}
	= \lin{ \nat -\bar\nat , \mu - \bar\mu} \; .
}
Now notice that~\eqref{eq:hanzely} features $\mu := \hat \mu_{n+1} = \hat\m_{n} - \lr g(\hat \nat_{n})$ stochastic and $\bar\mu := \bar\mu_{n+1} =  \hat\m_{n}   -\lr \nabla f(\hat \nat_{n})$ deterministic such that $\E_g[\mu] = \bar\mu$.
This means that
\alignn{
	\expect{\cS_\conj (\mu , \E[\mu])} 
	= \expect{\lin{\nat, \mu - \E[\mu]}}
	= \underbrace{\expect{\lin{\nat, \mu}} -\lin{\expect{\nat}, \expect{\mu} } }_{\Cov(\nat, \mu)}
	=\expect{\lin{\nat - \E[\nat],\mu}} 
	= \expect{\cS_\logpart(\nat, \E[\nat])} \; .
}
where the last equality holds by symmetry between the roles of $\logpart$ and $\conj$.
Note that the middle covariance formulation is actually the one  used by \citet{hanzely2018fastest}.
Now, \cref{eq:bias-variance} defines the variance as 
\alignn{
	\expect[1:n]{\bregmanconj(\tilde \m_n , \hat\m_n) }
	=\expect[1:n]{\bregman(\hat\nat_n , \E_{1:n}[\hat\nat_n]) }
	\leq \expect[1:n]{\cS_A(\hat\nat_n , \E_{1:n}[\hat\nat_n]) }
	=  \expect[1:n]{\cS_\conj(\hat\mu_n , \E_{1:n}[\hat\mu_n]) }\;,
	\label{eq:bregman-covariance}
}
where expectations $\E_{1:n}$ are on all samples $X_1, \dots, X_n$, whereas~\eqref{eq:hanzely} is written with 
\alignn{
	\expect[n]{\cS_\conj(\hat\mu_n , \E_{n}[\hat\mu_n]) } \leq \lr^2 C \; ,
}
where the expectation $\E_{n}$ is taken over only the last sample $X_n$, and the bound should hold uniformly over all $\hat\mu_{n-1}$.
Taking the expectation over $\hat\mu_{n-1}$ instead yields
\alignn{
	\expect[1:n]{\cS_\conj(\hat\mu_n , \E_{n}[\hat\mu_n]) }  \leq \lr^2 C \;.
}
The only difference with the RHS of \cref{eq:bregman-covariance} is in the inner expectation.
To overcome this difference, we need to plug in the form of $\hat \mu_n = \frac{n_0\mu_0 + \sum_i T_i}{n_0 +n}$. Notice that 
\alignn{
	&\MAPm - \E_n[\MAPm] = \frac{T_n - \m^*}{n_0+n} \\
	\implies & \expect[1:n]{\cS_\conj(\hat\mu_n , \E_{n}[\hat\mu_n]) }
	= \inv{n_0+n} \expect[1:n]{ \lin{\MAPt , T_1 - \m^*} }
}
while
\alignn{
	&\MAPm - \E_{1:n}[\MAPm] = \frac{\smallsum_i (T_i - \m^*)}{n_0+n} \\
	\implies & \expect[1:n]{\cS_\conj(\hat\mu_n , \E_{1:n}[\hat\mu_n]) }
	=\frac{n}{n_0+n} \expect[1:n]{ \lin{\MAPt , T_1 - \m^*} } \; .
}
In the end we get that the variance is dominated by $n$ times the expectation of \cref{eq:hanzely}
\alignn{
	\expect[1:n]{\bregmanconj(\tilde \m_n , \hat\m_n) }
	\leq n \expect[1:n]{\cS_\conj(\hat\mu_n , \E_{n}[\hat\mu_n]) }  \; .	
}
If assumption~\eqref{eq:hanzely} holds, then
\alignn{
\expect[1:n]{\bregmanconj(\tilde \m_n , \hat\m_n) }
\leq n \lr_n^2 C \in O\paren{\inv{n}} \; ,
}
where we assumed $\lr_n  \in O\paren{\inv{n}}$.
In conclusion, assuming~\eqref{eq:hanzely} holds uniformly or in expectation immediately implies a $O\paren{\inv{n}}$ convergence rate on the variance.
